A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-22 main.tex
example udpate
Ditch the Verma implementation
Closes: #34
Closes: #2

\section{Preliminaries}\label{sec:preliminaries}

\subsection{Monoids}

A monoid $\Monoid = \Tuple{X;\MonoidOp;0_{\Monoid}}$ is an algebraic structure
consisting of the non-empty carrier set of elements, $X$, an associative binary operation
$X \times X \rightarrow X$ denoted as $\MonoidOp$, that is where for all $a, b,
c \in X$, $(a \MonoidOp b) \MonoidOp c = a \MonoidOp (b \MonoidOp c)$. Finally,
$\Monoid$ must have an identity element $0_{\Monoid} \in X$ such that
$0_{\Monoid} \MonoidOp a = a \MonoidOp 0_{\Monoid} =   a$ for every $a \in X$.
We sometimes use integer multiplication to represent repeated application of
$\MonoidOp$, e.g. $3a = a \MonoidOp a \MonoidOp a$, for $a \in X$. $\Monoid$ is
called \textit{commutative} if $\MonoidOp$ also commutes, that is if $a
\MonoidOp b = b \MonoidOp a$ for all $a, b \in X$.

Finally, a \textit{homomorphism} is a structure-preserving map between two monoids
$\Monoid_1, \Monoid_2$, that is a map $\Map : S_1 \rightarrow S_2$ such that $\Map(a
\MonoidOp_1 b) = \Map(a) \MonoidOp_2 \Map(b)$, where $S_1, S_2$ are the carrier
sets of $\Monoid_1$ and $\Monoid_2$ respectively, and $\MonoidOp_1, \MonoidOp_2$
their binary operations.

\subsection{Languages, Finite-state Automata and their Products}

We define an alphabet as a finite set of symbols $\Alphabet$ with words $\Strings$, and
the concatenation operation as $s_1 \Concat{} s_2$ over two strings $s_1, s_2$.
Note that $\Strings = \Tuple{\Alphabet;\Concat{};\epsilon}$, is a
non-commutative monoid, referred to as the free monoid. The string length
function, $\Length{s}$ is an example of a homomorphism between $\Strings$
and~$\mathbb{Z}$.

An automaton~$\Automaton$ with alphabet~$\Alphabet$ is
$\AutomatonTuple$ where $\Transitions \subseteq \States \times \Alphabet \times
\States$, $\States$ is its states, $\InitialState$ its
initial state, and $\AcceptingStates$ is its set of accepting states.  We
write a transition $\Transition = \Tuple{\State, \Label, \State'} \in
\Transitions$ as $\Transition = \FromLabelTo{\State}{\Label}{\State'}$.
Similarly, we use the notation $\FromLabelTo{\State}{}{}$ to refer to the set of
transitions starting in $\State$, and $\FromLabelTo{}{}{\State}$ to refer to the
set of transitions coming into $\State$, whenever the automaton is clear from
the context.

We will let variables $\Transition, \Transition', \Transition_1, \ldots,
\Transition_n$ etc describe transitions, $\State, \ldots, \State_n$ states, and
$\Automaton, \ldots, \Automaton_n$ automata, and use subscript indexing
($\Transitions_\Automaton$) to refer to the transitions, states, etc of a given
automaton.

By a \emph{product} of two automata $\Automaton_1, \Automaton_2$, written
$\Automaton_1 \times \Automaton_2$, we mean an automaton constructed to run
$\Automaton_1$ and $\Automaton_2$ in parallel on an input and only accept the
input if both automata would do so.

We refer to the resulting product states as tuples, $\Tuple{\State, \State'}$,
which represent the state of the product automaton where $\Automaton_1$ would be
in $\State$ and $\Automaton_2$ would be in $\State'$. Note that since we use
ordered tuples the product is technically (but w.l.o.g) not commutative; the
left-hand-side must come from the left-hand term. The sole purpose of this
matching is to allow us to speak with precision about the origins of components
in a product.

\subsection{The Parikh Map and its Image}
Formally, the \textit{Parikh map} over an alphabet $\Alphabet=
\left\{a_1, \ldots, a_k \right\}$ is defined as in \cite{kozen}:
$$
\begin{aligned}
& \ParikhMap: \MapFromTo{\Strings}{\natural^k} \\
& \ParikhMap(s) = \VectorLiteral{\#a_1(s), \#a_2(s), \ldots, \#a_k(s)}
\end{aligned}
$$

That is, $\ParikhMap(s)$ is a vector of the number of occurrences of each
character in the language for a given string $s$. For example, for  $\Alphabet =
\Set{a, b}$, we would have $\ParikhMap(abb) = \VectorLiteral{1, 2}$.

We define the image of this map, the \textit{Parikh image}, of some subset of
the language $\Language \subseteq \Strings$ as:
\[
\ParikhMap(\Language) = \Set{\ParikhMap(x) \SuchThat x \in \Language}
\]

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$. We also sometimes use the standard
notation $\#l(w)$ to talk about an individual letter $l$ in a word $w$. For
example, for the Parikh vector above, we would have $\CountOf{a} = 1$. You have
already seen this in \cref{sec:introduction:parikh}.

Parikh's theorem states that any context-free language has a letter-equivalent
regular language (c.f.~\cite{construction} for a construction of such automata
from context-free grammars and~\cite{bounds} for bounds on its size). However,
there are languages that are not context-free that also have semilinear images
under~$\ParikhMap$ (e.g. $\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) =
\ParikhMap((abc)^*) = \CountOf{a} = \CountOf{b} = \CountOf{c} \land \CountOf{a}
\geq 0$). This means they can be represented as a quantifier-free Presburger
formula.

Note that while Parikh's theorem applies to context-free languages, in this
paper we focus only on regular languages.

\subsection{The Parikh image of a regular language expressed in Presburger arithmetic}
\label{sec:verma}

Since Parikh images are semilinear, any Parikh image can be written as a set of linear
equations. The following construction for an NFA $\Automaton = \AutomatonTuple$ is based on the version for context-free grammars presented in \cite{generate-parikh-image}:
\begin{equation}
\begin{aligned}
\ParikhMap(\Automaton) := 
& \AndComp{\Letter \in \Alphabet}{\LetterVar_{\Letter} = \sum_{\Transition = \FromLabelTo{}{\Letter}{} \in \Transitions} \TransitionVar_{\Transition}
}\\
& \AndComp{\State \in \AcceptingStates}{\FinalStateVar_{\State} > 0 \longrightarrow  \StateVar_{\State} > 0}\\
& \AndComp{\State \in \States}{
  \left(\FinalStateVar_{\State} \text{ if } \State \in \AcceptingStates \text{ otherwise } 0 \right) +
  \sum_{\Transition \in \FromLabelTo{\State}{}{}} \Filter(\Transition) - \sum_{\Transition \in \FromLabelTo{}{}{\State}} \Filter(\Transition)
= \begin{cases}
    1 \text{  if $\State = \InitialState$} \\
    0 \text{ otherwise}
  \end{cases}
}\\
& \AndComp{\Transition = \FromLabelTo{\State'}{}{\State} \in \Transitions}{
  \TransitionVar_{\Transition} > 0 \longrightarrow \StateVar_{\State} > 0
} \\
& \AndComp{\State \in F}{
  \StateVar_{\State} > 0 \longrightarrow \FinalStateVar_{\State} = 1
} \\
&\AndComp{\State \in \States}{
  \StateVar_{\State} > 0 \longrightarrow
  \OrComp{\Transition = \FromLabelTo{\State}{\Label}{\State'} \in \Transitions}{
    \StateVar_{\State} = \StateVar_{\State'} + 1 \land 
    \TransitionVar_{\Transition} \geq 1 \land
  \StateVar_{\State'} \geq 1
    }
}
\end{aligned}
\label{eq:generate-parikh}
\end{equation}

All variables $\TransitionVar_\Transition, \StateVar_\State, \FinalStateVar_\State$ are
existentially quantified and the free variables $\LetterVar_\Letter$ make up the
image. $\StateVar_\State$ represents the distance of state $\State$ from
$\InitialState$ in a spanning tree, $\TransitionVar_\Transition$ how many times
a transition $\Transition$ is used, and $\FinalStateVar_\State$ whether a given
accepting state $\State \in \AcceptingStates$ is the actually used final state.

In this paper we refer to this model as the baseline approach, though we also
apply optimisations as described in \cref{sec:implementing-baseline}. The
calculus introduced in this paper, by contrast, lazily enforces the
connectedness constraint of this encoding (the final three clauses) while also
interleaving the computation of products of automata and propagating information
between the steps to reduce the amount of work that needs to be done.

\subsection{Theorem Proving}

We assume basic familiarity with first-order logic, Presburger
arithmetic, and the classical sequent calculus. For reference, see
e.g. \cite{Fitting96a}.

@@

We start by defining our calculus, \Calculus{}, for one automaton, and only
extend it to products of automata in \cref{sec:multiple}. Assume an NFA
$\Automaton = \AutomatonTuple$ with $\NrTransitions$ transitions $\Transitions =
\Set{\EllipsisSequence{\Transition}{\NrTransitions}}$. For convenience, we then
introduce the following supporting notations:

@@

The greatest threat to validity is our choice of implementation platform and
automata library. There are some signs that product computation is inefficient,
notably the good performance of low-threshold automata materialisation that
prioritises computing smaller products. This means that \Calculus{} makes less
heavy use of product computation than baseline does, instead relying more on
\Princess{}'. This situation would unfairly advantage \Calculus{} if our
automata library was the bottleneck due to our automata library being
unnaturally poor. We believe this is unlikely since similar performance issues
have been reported for string solvers. Additionally, profiling suggests that
both \Princess{}-based backends spend most of their time in \Princess{},
suggesting that the automaton implementation is not the bottleneck.

>>>>

\section{Preliminaries}\label{sec:preliminaries}
\input{preliminaries.tex}

@@

We start by defining our calculus, \Calculus{}, for one automaton, and only
extend it to products of automata in \cref{sec:multiple}. We assume basic
familiarity with first-order logic, Presburger arithmetic, and the classical
sequent calculus. For reference, see e.g. \cite{Fitting96a}.

First, assume an NFA $\Automaton = \AutomatonTuple$ with $\NrTransitions$
transitions $\Transitions =
\Set{\EllipsisSequence{\Transition}{\NrTransitions}}$. For convenience, we then
introduce the following supporting notations:

@@

The greatest threat to validity is our choice of implementation platform and
automata library. There are some signs that product computation is inefficient,
notably the good performance of low-threshold automata materialisation that
prioritises computing smaller products. This means that \Calculus{} makes less
heavy use of product computation than baseline does, instead relying more on
\Princess{}. This situation would unfairly advantage \Calculus{} if our
automata library was the bottleneck due to our automata library being
unnaturally poor. We believe this is unlikely since similar performance issues
have been reported for string solvers. Additionally, profiling suggests that
both \Princess{}-based backends spend most of their time in \Princess{},
suggesting that the automaton implementation is not the bottleneck.
